### 2018-08-24 paper/preliminaries.tex
section 2.1 and additional explanations

@@

\subsubsection*{Communicating Sequential Processes} @todo
\fxwarning{alcc: I can make this small contribution.}

Throughout this paper, the alphabet of CSP processes $P$ is denoted by $\Sigma$.
Since we are only considering ``classical'' CSP processes, the alphabet is always finite.

\begin{itemize}
  \item traces and failures model
  \item refinement and equivalence
\end{itemize}

The FDR tool~\cite{fdr} supports model checking and semantic analyses of CSP
processes.

@@

\begin{example}\label{ex:a}
Consider the CSP process $P$ defined below, and $\Sigma = \{a,b,c\}$.
\begin{eqnarray*}
P & = & a \then (Q\intchoice R)
\\
Q & = & a \then P \extchoice c \then P
\\
R & = & b \then P \extchoice c \then R
\end{eqnarray*}
Its transition graph $G(P)$ is shown in Fig.~\ref{fig:tga}. Process state
$P/\langle\rangle$ is represented there as Node\_0, with $\{ a\}$ as the
only minimal acceptance, since $a$ can never be refused, and no other events are
accepted. Having engaged into $a$, the transition emanating from Node\_0
leads to Node\_2 representing  the process state $P/a = Q\intchoice R$. The
internal choice operator induces several minimal acceptances derived from $Q$ and
$R$. Since these processes accept their initial events in external choice,
$Q\intchoice R$ induces minimal acceptance sets $\{a,c\}$ and
$\{b,c\}$. Note that event $c$ can never be refused, since it is contained in each minimal acceptance set.

@@

Refinement relations between finite-state CSP processes $P, Q$ can be be
expressed by means of their normalised transition graphs in the following
way. %\fixme{alcc: trc was defined for P, not G(P) and this lemma is used
%later for trc(P). So, we need to be clearer as to what we mean.}
%\fxnote{jp: I have now introduced notation $\trc(G(P))$ above, so I think that the
%statement of the lemma is now sufficiently clear.}
%
\begin{lemma}
  \label{lemma:tgtrcref}
  \begin{eqnarray}
  P \lessdet_T Q & \Leftrightarrow & \trc(G(Q)) \subseteq\trc(G(P))
  \label{eq:trcrefa}
  \\
  \label{eq:failrefa}
  P \lessdet_F Q & \Leftrightarrow & \trc(G(Q)) \subseteq\trc(G(P)) \wedge {} \nonumber
  \\ & & \forall s\in\trc(G(Q)), R_Q\in\maxrefs(G(Q)/s):  \nonumber
  \\ & & \tabd
  \exists R_P\in\maxrefs(G(P)/s): R_Q\subseteq R_P
  \\
  \label{eq:failrefb}
  P \lessdet_F Q & \Leftrightarrow & \trc(G(Q)) \subseteq\trc(G(P)) \wedge {} \nonumber
  \\ & & \forall s\in\trc(G(Q)), A_Q\in\minaccs(G(Q)/s): \nonumber
  \\ & & \tabd
  \exists A_P\in\minaccs(G(P)/s): A_P\subseteq A_Q
  \end{eqnarray}
  \xbox
\end{lemma}
\fixme{alcc: I'll write an explanation of these results.}

>>>>

In this section we present the background material for our work.

@@

First, we introduce the CSP notation and refinement notions. We then discuss
a graph-based semantics for finite-state CSP processes.

% =========================================================================
\subsubsection*{CSP}
\label{section:CSP}

CSP is a process algebra for refinement. Using CSP, we model both systems and
their components using processes. There are characterised by their patterns
of interactions, modelled by synchronous, instantaneous, and atomic events.

A prefixing operator $e \then P$ defines a process that is ready to engage in
the event $e$, pending agreement of its environment to synchronising. After
$e$ occurs, the process behaves as defined by $P$. The environment can be
other processes, in parallel, or the environment of a system as a whole.

Two forms of choice support branching behaviour. An external choice $P
\extchoice Q$ between processes $P$ and $Q$ offers to the environment the
initial events of $P$ and $Q$. Once synchronisation takes place, the process
that offered this event is chosen and the other is discarded. In an internal
choice $P \intchoice Q$, the environment does not have an opportunity to
interfere:~the choice is made by the process itself.

\begin{example}
  We consider the processes $P$, $Q$, and $R$ below. $P$ is initially
  ready to engage in the event $a$, and then makes a choice to
  behave as $Q$ or $R$.
  \begin{eqnarray*}
  P & = & a \then (Q\intchoice R)
  \\
  Q & = & a \then P \extchoice c \then P
  \\
  R & = & b \then P \extchoice c \then R
  \end{eqnarray*}
  $Q$, for instance, offers to the environment the choice to engage in $a$
  again or $c$. In both cases, afterwards, we have a recursion back to $P$.
  In $R$, if $b$ is chosen, we also have a recursion back to $P$. If $c$ is
  chosen, the recursion is to $R$.
  \xbox
\end{example}
%
Iterated forms $\Extchoice i: I @ P(i)$ and $\Intchoice i: I @ P(i)$ of the
external and internal choice operators allow us to define a choice over a
collection of processes $P(i)$.

There are several parallelism operators. A widely used form of parallelism $P
\parallel[cs] Q$ defines a process in which the behaviour is characterised by
those of $P$ and $Q$ in parallel, synchronising on the events in the set
$cs$. Other forms of parallelism can be defined using this operator.

Interactions that are not supposed to be visible to the environment can be
hidden. The operator $P \ H$ defines a process that behaves as $P$, with the
interactions modelled by events in the set $H$ hidden. Frequently, hiding in
used in conjunction with parallelism, so that communications in a network of
parallel processes used just to coordinate them are affected
by the environment.

A rich collection of process operators allow us to define networks of
parallel processes in a concise and elegant way, and reason about safety,
liveness, and divergences.  A comprehensive account of the notation is given
in~\cite{Roscoe2010}.

A distinctive feature of CSP is its treatment of refinement~(as opposed to
bisimulation), which is convenient for reasoning about program correctness,
due to its treatment of nondeterminism and divergence.  A variety of semantic
models capture different notions of refinement. The simplest model
characterises a process by its possible traces:~sequence of events in which
it can engage.  We say that a process $P$ is trace-refined by another process
$Q$, written $P \lessdet_T Q$, if the set $traces(Q)$ of traces of $Q$ is
contained in the set $traces(P)$ of traces of $P$.

In fact, in every model, subset containment is used to define refinement. The
model we focus here first is the failures model, which captures both
sequences of interactions and deadlock behaviour. A failure of a process $P$
is a pair $(t,X)$ containing a trace $t$ of $P$ and a refusal:~a set $X$ of
events in which $P$ may refuse to engage, having engaged in the events of
$t$. The failures model of a process $P$ records all its failures in a set
$failures(P)$.

Failures refinement, $\lessdet_F$, is defined by set containment, but can
also be expressed in terms of traces refinement and another conformance
relation called $conf$ and defined below. The set $Ref(P,t)$ used in this
definition, and also specified below, contains the refusals of $P$ after the
trace $t$.
%
\begin{displaymath}
  Q\ conf\ P \defs \forall t: traces(P) \cap traces(Q) @ Ref(Q,t)
  \subseteq Ref(P,t)
  \\ %
  \mbox{where } Ref(P,t) \defs \{~X | (t,X) \in failures(P)~\}
\end{displaymath}
%
Conformance according to $conf$ requires that no new deadlocks are introduced
in an implementation. The theorem below, proved in~\cite{CG07}, establishes
that failures refinement amounts to traces refinement and $conf$.
%
\begin{theorem}
  $(P \lessdet_F Q) \iff (P \lessdet_T Q \land Q\ conf\ P)$
  \label{thm:fref-tref-conf}
\end{theorem}
%
Process equivalence in each model is given simply by mutual refinement.

Throughout this paper, the alphabet of the processes, that is the set of
events that are in scope, is denoted by $\Sigma$. Since we are only
considering ``classical'' CSP processes, the alphabet is always finite. The
FDR tool~\cite{fdr} supports model checking and semantic analyses of
finite-state CSP processes.

@@

\begin{example}\label{ex:a}
Consider the CSP process $P$ in Example~\ref{example:CSP}. Its transition
graph $G(P)$ is shown in Fig.~\ref{fig:tga}. Process state $P/\langle\rangle$
is represented there as Node\_0, with $\{ a\}$ as the only minimal
acceptance, since $a$ can never be refused, and no other events are accepted.
Having engaged into $a$, the transition emanating from Node\_0 leads to
Node\_2 representing  the process state $P/a = Q\intchoice R$. The internal
choice operator induces several minimal acceptances derived from $Q$ and $R$.
Since these processes accept their initial events in external choice,
$Q\intchoice R$ induces minimal acceptance sets $\{a,c\}$ and $\{b,c\}$. Note
that event $c$ can never be refused, since it is contained in each minimal
acceptance set.

@@

Refinement relations between finite-state CSP processes $P, Q$ can be be
expressed by means of their normalised transition graphs in the following
way. %\fixme{alcc: trc was defined for P, not G(P) and this lemma is used
%later for trc(P). So, we need to be clearer as to what we mean.}
%\fxnote{jp: I have now introduced notation $\trc(G(P))$ above, so I think that the
%statement of the lemma is now sufficiently clear.}
%
\begin{lemma}
  \label{lemma:tgtrcref}
  \begin{eqnarray}
  P \lessdet_T Q & \Leftrightarrow & \trc(G(Q)) \subseteq\trc(G(P))
  \label{eq:trcrefa}
  \\
  \label{eq:failrefa}
  P \lessdet_F Q & \Leftrightarrow & \trc(G(Q)) \subseteq\trc(G(P)) \wedge {} \nonumber
  \\ & & \forall s\in\trc(G(Q)), R_Q\in\maxrefs(G(Q)/s):  \nonumber
  \\ & & \tabd
  \exists R_P\in\maxrefs(G(P)/s): R_Q\subseteq R_P
  \\
  \label{eq:failrefb}
  P \lessdet_F Q & \Leftrightarrow & \trc(G(Q)) \subseteq\trc(G(P)) \wedge {} \nonumber
  \\ & & \forall s\in\trc(G(Q)), A_Q\in\minaccs(G(Q)/s): \nonumber
  \\ & & \tabd
  \exists A_P\in\minaccs(G(P)/s): A_P\subseteq A_Q
  \end{eqnarray}
  \xbox
\end{lemma}
%
Traces refinement is, as already explained, characterised by subset inclusion
of sets of traces, here defined in terms of the normalised graph of a
process. In~(\ref{eq:failrefa}), we have a result similar to
Theorem~\ref{thm:fref-tref-conf}, but rely on the fact that refusal sets are
subset closed. In other words, if a process may refuse to engage in all
events in a set $X$, then it can refuse to engage in all events of subsets of
$X$. So, in considering $conf$, it is enough to consider maximal refusals.
Finally, (\ref{eq:failrefb}) captures the same result using acceptances,
rather than refusals, which as already observed, can be calculated by taking
the complement of refusals. So, acceptances of $Q$ are required to have a
corresponding larger set of acceptances in $P$.
